# Installation


## Install from GitHub
All epiregulon components are available on [GitHub](https://github.com/xiaosaiyao?tab=repositories). There are three separate packages: <br>
- `epiregulon` constructs the GRN. <br>
- `epiregulon.extra` provides tools for visualization, differential TF analysis and network analysis. <br>
- `epiregulon.archr` provides integration with `ArchR`. <br> 

```{r installation, message=FALSE, eval=FALSE}
# install devtools
if(!require(devtools)) install.packages("devtools")

devtools::install_github(repo='xiaosaiyao/epiregulon')
devtools::install_github(repo='xiaosaiyao/epiregulon.extra')
devtools::install_github(repo='xiaosaiyao/epiregulon.archr')

```

The data package `scMultiome` stores ChIP-seq data and pre-processed scMultiome datasets.

```{r message=FALSE, eval=FALSE}
devtools::install_github(repo='xiaosaiyao/scMultiome')
```

## Install from Bioconductor

Epiregulon is now available through [Bioconductor](https://www.bioconductor.org/).

```{r, results='hide', message=FALSE, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
 
BiocManager::install("epiregulon")
BiocManager::install("epiregulon.extra")
BiocManager::install("scMultiome")

```

