# Installation


## Installing from github
All the epiregulon components are available on [github](https://github.com/xiaosaiyao?tab=repositories). 
There are three separate epiregulon packages. The core epiregulon package supports input in the form of `SingleCellExperiment` objects. If the users would like to start from `ArchR` projects, they may choose to use `epiregulon.archr` package, which allows for the seamless integration with `ArchR` package through accepting its output to be used in the downstream workflow. 

```{r installation, message=FALSE, eval=FALSE}
# install devtools
if(!require(devtools)) install.packages("devtools")

# install basic epiregulon package
devtools::install_github(repo='xiaosaiyao/epiregulon')

# install extended version of epiregulon
devtools::install_github(repo='xiaosaiyao/epiregulon.archr')
```
 
Moreover, we provide a suite of tools for the enrichment analysis, visualization, and network analysis which can be run on the `epireglon` or `epiregulon.archr` output.

```{r message=FALSE, eval=FALSE}
# install extended version of epiregulon
devtools::install_github(repo='xiaosaiyao/epiregulon.extra')
```

The data package scMultiome is for storing TF ChIP-seq data and pre-processed datasets.

```{r message=FALSE, eval=FALSE}
# install extended version of epiregulon
devtools::install_github(repo='xiaosaiyao/scMultiome')
```

## Installing from Bioconductor

Epiregulon is now available through [Bioconductor]("https://www.bioconductor.org/).

```{r, results='hide', message=FALSE, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
 
BiocManager::install("epiregulon")
BiocManager::install("epiregulon.extra")
BiocManager::install("scMultiome")

```

## Citation

Check out our preprint

Epiregulon: Inference of single-cell transcription factor activity to dissect mechanisms of lineage plasticity and drug response
Tomasz Włodarczyk, Aaron Lun, Diana Wu, Shreya Menon, Shushan Toneyan, Kerstin Seidel, Liang Wang, Jenille Tan, Shang-Yang Chen, Timothy Keyes, Aleksander Chlebowski, Yu Guo, Ciara Metcalfe, Marc Hafner, Christian W. Siebel, M. Ryan Corces, Robert Yauch, Shiqi Xie, Xiaosai Yao
bioRxiv 2023.11.27.568955; doi: [https://doi.org/10.1101/2023.11.27.568955]("https://www.biorxiv.org/content/10.1101/2023.11.27.568955v3")
