```{r echo=FALSE}
library(BiocStyle)
```
# Quick start

In this chapter, we illustrate the epiregulon workflow, using the reprogram-seq dataset which can be downloaded from the `r Biocpkg("scMultiome")` package. In this example, prostate cancer cells (LNCaP) were infected in separate wells with viruses encoding 4 transcription factors (NKX2-1, GATA6, FOXA1 and FOXA2) and a positive control (mNeonGreen) before pooling. The identity of the infected transcription factors was tracked through cell hashing (available in the field `hash_assignment` of the `colData`) and serves as the ground truth.

## Data preparation

Prior to using `epiregulon`, single cell preprocessing needs to be performed by user's favorite methods. The following components are required: <br>
1. Peak matrix from scATAC-seq containing the chromatin accessibility information <br>
2. Gene expression matrix from either paired or unpaired scRNA-seq. RNA-seq integration needs to be performed for unpaired dataset. <br>
3. Dimensionality reduction matrix from with either single modalities or joint scRNA-seq and scATAC-seq <br>

```{r, message=FALSE}
# load the MAE object
library(scMultiome)
library(epiregulon)

mae <- scMultiome::reprogramSeq()

# extract peak matrix
PeakMatrix <- mae[["PeakMatrix"]]

# extract expression matrix
GeneExpressionMatrix <- mae[["GeneExpressionMatrix"]]
rownames(GeneExpressionMatrix) <- rowData(GeneExpressionMatrix)$name

# define the order of hash_assigment
GeneExpressionMatrix$hash_assignment <- factor(as.character(GeneExpressionMatrix$hash_assignment),
                                               levels = c("HTO10_GATA6_UTR", "HTO2_GATA6_v2", "HTO8_NKX2.1_UTR", "HTO3_NKX2.1_v2", 
                                                          "HTO1_FOXA2_v2", "HTO4_mFOXA1_v2", "HTO6_hFOXA1_UTR", "HTO5_NeonG_v2"))
# extract dimensional reduction matrix
reducedDimMatrix <- reducedDim(mae[['TileMatrix500']], "LSI_ATAC")

# transfer UMAP_combined from TileMatrix to GeneExpressionMatrix
reducedDim(GeneExpressionMatrix, "UMAP_Combined") <- reducedDim(mae[['TileMatrix500']], "UMAP_Combined")
```
Visualize singleCellExperiment by UMAP

```{r}

scater::plotReducedDim(GeneExpressionMatrix, 
                       dimred = "UMAP_Combined", 
                       text_by = "Clusters", 
                       colour_by = "Clusters")
```

## Retrieve bulk TF ChIP-seq binding sites 

First, we retrieve a `GRangesList` object containing the binding sites of all the transcription factors and co-regulators. These binding sites are derived from bulk ChIP-seq data in the [ChIP-Atlas](https://chip-atlas.org/) and [ENCODE](https://www.encodeproject.org/) databases. For the same transcription factor, multiple ChIP-seq files from different cell lines or tissues are merged. For further information on how these peaks are derived, please refer to `?epiregulon::getTFMotifInfo`. Currently, human genomes hg19 and hg38 and mouse mm10 are supported. 

```{r getTFMotifInfo}
grl <- getTFMotifInfo(genome = "hg38")
grl
```

### Link ATAC-seq peaks to target genes

Next, we try to link ATAC-seq peaks to their putative target genes. We assign a peak to a gene within a size window (default ±250kb) if the chromatin accessibility of the peak and expression of the target genes are highly correlated (default threshold 0.5). To compute correlations, we first create cell aggregates by performing k-means clustering on the reduced dimensionality matrix. To estimate the optimal number of cells per cluster we will run the function `optimizeMetacellNumber`.

```{r}
set.seed(1010)
cellNum <- optimizeMetacellNumber(expMatrix = GeneExpressionMatrix,
                                  peakMatrix = PeakMatrix,
                                  exp_assay = "normalizedCounts",
                                  peak_assay = "counts",
                                  reducedDim = reducedDim(GeneExpressionMatrix, "UMAP_Combined"),
                                  BPPARAM = BiocParallel::SerialParam(progressbar = FALSE))

cellNum
```

`cellNum` object is now provided as an argument to `calculateP2G`.

```{r calculateP2G}
set.seed(1010)
p2g <- calculateP2G(peakMatrix = PeakMatrix, 
                    expMatrix = GeneExpressionMatrix, 
                    reducedDim = reducedDimMatrix,
                    exp_assay = "normalizedCounts",
                    peak_assay = "counts")

p2g
```

### Add TF motif binding to peaks

The next step is to add the TF binding information by overlapping regions of the peak matrix with the bulk chip-seq database. The output is a data frame object with three columns:

1. `idxATAC` - index of the peak in the peak matrix
2. `idxTF` - index in the gene expression matrix corresponding to the transcription factor
3. `tf` - name of the transcription factor


```{r addTFMotifInfo}
overlap <- addTFMotifInfo(grl = grl, p2g = p2g, peakMatrix = PeakMatrix)
head(overlap)
```

### Generate regulons

A data frame representing the inferred regulons is then generated. The data frame consists of several columns, including those indicating target genes, transcription factors, and regulatory elements indexed by their corresponding rows in the `PeakMatrix`.

```{r, warning=FALSE, getRegulon}
regulon <- getRegulon(p2g = p2g, overlap = overlap, aggregate = FALSE)
regulon
```

### Network pruning (recommended)

Since our regulon is quite large, and likely contains many false connections, we will pass it through the pruning function.

```{r network pruning, results = "hide", message = FALSE}

pruned.regulon <- pruneRegulon(expMatrix = GeneExpressionMatrix,
                               exp_assay = "normalizedCounts",
                               peakMatrix = PeakMatrix,
                               peak_assay = "counts",
                               test = "chi.sq",
                               regulon[regulon$tf %in% c("NKX2-1","GATA6","FOXA1","FOXA2", "AR"),],
                               clusters = GeneExpressionMatrix$Clusters,
                               prune_value = "pval",
                               regulon_cutoff = 0.05
                               )

pruned.regulon
```

### Calculate TF activity 

Finally, the activities for a specific TF in each cell are computed by averaging expressions of target genes linked to the TF weighted by the test statistics of choice, chosen from either correlation, mutual information or the Wilcoxon test statistics. 
$$y=\frac{1}{n}\sum_{i=1}^{n} x_i * weights_i$$
where $y$ is the activity of a TF for a cell,
$n$ is the total number of targets for a TF,
$x_i$ is the log count expression of target $i$ where $i$ in {1,2,...,n} and
$weights_i$ is the weight of TF - target $i$, calculated by running `addWeights` function.



```{r calculateActivity}
regulon.w <- addWeights(regulon = pruned.regulon,
                        expMatrix  = GeneExpressionMatrix,
                        exp_assay  = "normalizedCounts",
                        peakMatrix = PeakMatrix,
                        peak_assay = "counts",
                        clusters = GeneExpressionMatrix$Clusters,
                        method = "wilcox")


score.combine <- calculateActivity(expMatrix = GeneExpressionMatrix,
                                   regulon = regulon.w, 
                                   mode = "weight", 
                                   method = "weightedMean", 
                                   exp_assay = "normalizedCounts",
                                   normalize = FALSE)

score.combine[1:5,1:5]
```

## Session Info

```{r}
sessionInfo()
```
